### 2023-02-07 doc/manuscript/translocation.qmd
Make suggestions on wording in results to improve between section flow

<!--# Likely need an introductory few sentences to put the translocations in the proper context. -->During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap). Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap). Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival). As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details). The relatively constant within-site frog survival across all translocations suggests an important effect of site characteristics on frog survival. <!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)? -->

@@

@@

<!--# Mention the improved survival in recent compared with early translocations resulting from our better-informed site selection? Would put the viability results in better context because most of the low-viability populatons are those resulting from early translocations. -->

@@

<!--# Likely need an introductory few sentences to put the viability analyses in the proper context. -->Given observed yearly adult survival probabilities, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ near or greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 0.95-1.16 for these six populations). When the probability of successful recruitment $\omega$ was greater than 0.1 (the conservative estimate we used as a baseline value in our model), long-run growth rate of these populations could be notably greater than 1 (@fig-viability A). In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A). A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).

Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B). Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction. For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B). In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1. To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years). <!--# To help constrain the possible values of omega, add information about recruitment probability observed in the study populations? Because we don't have accurate counts of subadults, can we even provide rough estimate of recruitment probability? -->

@@

@@

![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon). To obscure the exact locations of populations, random noise was added to all point coordinates. Elevations are indicated as in @fig-exomemap. See inset map in @fig-exomemap to locate Yosemite National Park within California. <!--# Move to Supporting Information. Add YOSE boundary to inset map of fig-exomemap? Also, is basemap necessary/helpful? Finally, better to add color coding showing which donor was used at which sites?-->](figures/map_translocation_points.png){#fig-yosemap width="450"}

![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study. Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site. Frog survival was estimated using capture-mark-recapture methods. Dot colors indicate the donor population from which frogs in each translocated cohort were collected.](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}

@@

![<!--# Add first sentence that provides a brief overview of the entire figure. --> A. The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R. sierrae*. Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line). B. The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$. The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}

![<!--# Add first sentence that provides a brief overview of the entire figure. --> A. The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch). Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3. B. The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}

@@

![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R. sierrae* populations. The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests are role for RIN3 in frog resistance against Bd infection and associated population recovery. ](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}

@@

@@

To determine the implications of acquired Bd resistance on long-term population viability, we developed a dynamic, population model for *R. sierrae*. Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations? We developed a model that tracked seven state variables of an *R. sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$). We divided adults into two classes $A_T$ and $A_R$ because there was evidence that the survival probability of translocated adults and naturally recruited adults differed.

@@

In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction. We also did not directly consider the dynamics of Bd in this model. We made this decision because i) all translocated populations are infected with Bd at high prevalence, and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022]. Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.

@@

The baseline parameter values for the model and how they were estimated are given in Table 1. Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study. For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$. For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.

Table 1: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. <!--# Roland, please add any additional details needed to the "Source" column.  We ideally want to keep this succinct.  -->

@@

>>>>

<!--# Likely need an introductory few sentences to put the translocations in the proper context. -->

<!--# Mark W: Pulling from the methods, we say: "Translocations of [putatively resistant] *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species."  Could be a good starting sentence.  Also, as I am reading through the results, our sections do not elude to the Allie's section.  We should discuss how our sections can set-up Allie's section with a few well-chosen words throughout.  -->

During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of adult frogs to `r n_distinct(d1$site_id)` recipient sites in Yosemite (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap). Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap). Estimates of frog survival during the one year following translocation indicate that survival was highly variable between sites, but relatively constant within sites (for the subset of sites that received multiple translocations; @fig-translocation-survival). As such, for sites that received more than one translocation, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (median of posterior distribution = 4.14, uncertainty interval = 3.06\--5.28; see Supporting Information for analysis details). The relatively constant within-site frog survival across all translocations suggests an important effect of site characteristics on frog survival. <!--# How to extract these values from table produced by line 934 in translocation_survival_analysis.Rmd (or from associated code output)? -->

@@

<!--# Suggested text to wrap this section up and build into the next two sections: Overall, our analyses show that putatively resistant frogs can maintain the benefits of resistance (i.e., minimal Bd-induced mortality despite ongoing infection) in non-natal habitats. Moreover, in three locations where longer mark-recapture time series allow us to assess the survival of adults recruited to the population due to the breeding of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults. These results show that putative resistance in non-natal habitats is maintained across generations of frogs, suggesting that acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations.  We test these claims in the following sections. -->

<!--# Mark W: Regarding the above paragraph, what evidence do we have of "tolerance"? Are Bd loads significantly reduced in populations compared to epizootic levels? Can we convincingly state that the likely mechanism is more consistent with resistance or tolerance?  I ask because it would be nice to just use one of these terms in the paper and not get caught up in any resistance vs. tolerance discussions with reviewers.  -->

@@

<!--# Mention the improved survival in recent compared with early translocations resulting from our better-informed site selection? Would put the viability results in better context because most of the low-viability populatons are those resulting from early translocations. Mark W: Yes, I think this is important.  I think we want to drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd. -->

@@

<!--# Likely need an introductory few sentences to put the viability analyses in the proper context.  Mark W: I have made an attempt at this. -->

To test whether the observed yearly survival probabilities of putatively resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity. We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values). Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations). When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A). In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A). A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).

Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B). Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction. For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B). In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1. To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years). Overall, our model demonstrates that given observed yearly adult survival probabilities of putatively resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.

<!--# To help constrain the possible values of omega, add information about recruitment probability observed in the study populations? Because we don't have accurate counts of subadults, can we even provide rough estimate of recruitment probability?  Mark W here: This would be great. If we have even ballpark estimates of \omega it would be useful to inclue.-->

@@

#### Conclusions

<!--# Placeholder for our conclusions section -->

@@

![Map showing the locations of translocated and donor populations in Yosemite National Park (park boundary indicated by black polygon). To obscure the exact locations of populations, a small amount of random noise was added to all point coordinates. Elevations are indicated as in @fig-exomemap. See inset map in @fig-exomemap to locate Yosemite National Park within California. <!--# Move to Supporting Information. Add YOSE boundary to inset map of fig-exomemap? Also, is basemap necessary/helpful? Finally, better to add color coding showing which donor was used at which sites?  Mark W: I like the basemap.  -->](figures/map_translocation_points.png){#fig-yosemap width="450"}

![Median survival of each cohort of translocated frogs to the 12 recipient sites included in this study. Sites are arranged along the x-axis using the average of the median 1-year survival per translocation at each site. Frog survival was estimated using capture-mark-recapture methods. Dot colors indicate the donor population from which frogs in each translocated cohort were collected.<!--# Mark W: Do we want to include error bars on these median estimates?  -->](figures/translocation_survival_bysiteid.png){#fig-translocation-survival width="450"}

@@

![A population viability analysis of translocated *R. sierrae* that quantifies the long-run growth rate and extinction probability of frog populations. A. The predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model for translocated *R. sierrae*. Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (given by the dashed line). B. The predicted fifty-year extinction probabilities of the twelve translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$. The dashed line shows where $\omega = 0.3$, for comparison to A.](figures/pop_viability_figures_for_manuscript_v2.jpg){#fig-viability}

![Sensitivity analysis and empirical validation of the stage-structured *R. sierrae* model. A. The elasticity of $\lambda$ with changes in three parameters: $\sigma_{A_R}$, $\omega$, and *F (*number of eggs produced by a female frog in a year that successfully hatch). Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3\$. B. The 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp}

@@

![Allele frequencies of the outlier SNPs for each of the sampled naive and recovering *R. sierrae* populations. The results for RIN3 are consistent with strong selection on this gene following the arrival of Bd, and suggests are role for RIN3 in frog resistance against Bd infection and associated population recovery.](figures/outlier_map_draft_figure.png){#fig-allelefrequencies}

@@

<!-- Mark W: We should specify whether the default priors were uninformative, slightly informative, etc. -->

@@

To determine the implications of observed yearly adult survival on the long-term population viability of putatively resistant frogs in non-natal habitats, we developed a population model for *R. sierrae*. Our central question was: how does the magnitude and variation in observed adult survival probability across translocated populations affect the long-term persistence probability of populations? We developed a model that tracked seven state variables of an *R. sierrae* population: the density of translocated adults ($A_T$), the density of adults naturally recruited into the population ($A_R$), the density of first year larvae ($L_1$), the density of second year larvae ($L_2$), the density of third year larvae ($L_3$), the density of first year juveniles ($J_1$), and the density of second year juveniles ($J_2$). We divided adults into two classes $A_T$ and $A_R$ because there was evidence that the survival probability of translocated adults and naturally recruited adults differed.

@@

In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction. We also did not directly consider the dynamics of Bd in this model. We made this decision because i) translocated populations are infected with Bd at high prevalence [@joseph2018], and ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022]. Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.

@@

The baseline parameter values for the model and how they were estimated are given in @tbl-param_values. Parameters $\sigma_{A_T}$ and $\sigma_{A_R}$ were extracted directly from our CMR model described previously in this study. For populations where we had a sufficient number of pit-tagged, naturally recruited adults, we observed that $\sigma_{A_T}$ and $\sigma_{A_R}$ could be notably different, with $\sigma_{A_R} > \sigma_{A_T}$. For populations lacking any naturally-recruited adults, we were unable to directly estimate $\sigma_{A_R}$, and instead set $\sigma_{A_R} = \sigma_{A_T}$.

@@

: Description and values of parameters used in the model. All survival probabilities are in the presence of the fungal pathogen Bd. {#tbl-param_values}
